Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-20 pkg/disttree/inst/draft/draft.Rnw
merged conference version of abstract into manuscript draft, some smaller touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2063 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9996ea3546043b2e9c1118daa67c0fd67c538abe

\Abstract{
While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to model all distribution parameters. In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting. In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) 
each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects. A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{Loh:2011}) 
or random forests. %(\citealt{Breiman:2001}). 
%%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
This class of tree and forest models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods. In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. 
%%%<TH> break-up very long sentence </TH>
Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}

@@

In regression analysis a wide range of models has been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
    encompassing an additional nonlinear link function for the conditional mean.
  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item \emph{Generalized additive models for location, scale, and shape}
    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

\subsection{Motivation}
As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed 
in the field of parametric modeling. Especially the shift from conditional mean 
regression models to distributional regression models has increased the application 
spectrum by allowing for much broader inference. In some cases further features 
such as the ability to capture non-additive effects or interactions are desirable 
which demands for more flexibility. Since tree-structured models provide these 
required features, combining them with parametric models can offer a framework 
that is able to deal with a wide variety of different modeling situations. 
In this section the novel methodology which embeds parametric modeling into 
the idea of tree-structured models will be introduced step by step. Starting 
with a simple distributional fit in Section~\ref{distfit} will be followed by 
building a distributional tree in Section~\ref{disttree} and finally a distributional 
forest in Section~\ref{distforest}.

@@

The goal is to fit a distributional model $D(Y, \theta)$ to a response variable 
$Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters 
$\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution 
family with density function $f$ has to be specified in advance such that a 
log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided. Using 
the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions 
can be selected where each distribution parameter can be specified separately. 
This enables the model to deal with various features that can make the task of 
fitting a model very challenging such as censoring, truncation or heavy tails.

@@

A common way of including these covariates is to consider them as regressors 
in a distributional model such that each parameter $\theta$ depends on covariates 
$\bold{Z} = (Z_1,\ldots,Z_m)$. This could for example be done by a linear function 
$g$. In that way the fitted model would be specified as 
$D(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ 
to specify a more flexible model 
$D(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
Another approach to distributional regression is proposed by 
\cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex 
distribution can be modeled by an additive term consisting of a variety of 
different functional effect types. These include non-linear effects, spatial 
effects, random coefficients and interaction surfaces which extends the 
flexibility compared to parametric models such as GAMLSS modeling only 
smooth effects.

@@

\subsection{Distributional tree}
\label{disttree}
Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$, 
the idea is to use the information provided by this model together with a 
set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ 
to decide whether the data set should be split into subgroups or not. 
In particular, the information provided by the fitted model are the scores 
$$
s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
$$ 
These values express how well the model fits the data regarding each 
parameter. Ideally they should fluctuate randomly around zero, 
similar to residuals in ordinary least squares estimation, but in 
this case there is a score value for each pair of estimated parameter 
$\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ 
and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$. This enables the 
model to be sensitive to changes in each of the parameters which is one 
of the main advantages of distributional trees over other tree models 
considering only the mean of the distribution.

@@

\begin{itemize}
\item A distributional model $D(Y, \theta)$ is fit to the set of 
observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
parameter $\hat{\theta}$. This parameter can be a multidimensional 
vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
depending on the type of distribution. The estimation is done by 
maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
which is equal to solving
$$
\sum_{i=1}^n s(\theta, y_i) = \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\theta; y_i) = 0
$$
for $\theta$.

>>>>

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameter(s) of the distribution are linked to regressors. In many classical
models this often only captures the location/expectation of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression models both
mean and variance of a Gaussian response variable and is particularly
popular in weather forecasting and, more generally, the GAMLSS framework
allows to establish generalized additive models for location, scale, and shape
with smooth linear or nonlinear effects.
%
However, in situations where variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous locations in a mountainous
region (Tyrol, Austria) based on a very large number of numerical weather
prediction quantities. Comparisons with GAMLSS using either a specification
based on prior meteorological knowledge or a computationally more demanding
boosting approach show that distributional random forests automatically
select variables and interactions, performing on par or often even better than
the GAMLSS approaches.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}

@@

In regression analysis a wide range of models has been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:

\pagebreak

\begin{itemize}
  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
    encompassing an additional nonlinear link function for the conditional mean.
  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item \emph{Generalized additive models for location, scale, and shape}
    (GAMLSS, \citealt{Rigby+Stasinopoulos:2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

As illustrated in Figure~\ref{Devel_parmod} a wide range of tools has been developed 
in the field of parametric modeling. Especially the shift from conditional mean 
regression models to distributional regression models has increased the application 
spectrum by allowing for much broader inference. In some cases further features 
such as the ability to capture non-additive effects or interactions are desirable 
which demands for more flexibility. Since tree-structured models provide these 
required features, combining them with parametric models can offer a framework 
that is able to deal with a wide variety of different modeling situations. 
In this section the novel methodology which embeds parametric modeling into 
the idea of tree-structured models will be introduced step by step. Starting 
with a simple distributional fit in Section~\ref{distfit} will be followed by 
building a distributional tree in Section~\ref{disttree} and finally a distributional 
forest in Section~\ref{distforest}.

@@

The goal is to fit a distributional model $\mathcal{D}(Y, \theta)$ to a response variable 
$Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters 
$\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution 
family with density function $f$ has to be specified in advance such that a 
log-likelihood function $\ell(\theta; Y) = \log(f(Y; \theta)$ is provided. Using 
the GAMLSS framework (\citealt{Rigby+Stasinopoulos:2005}) a wide range of distributions 
can be selected where each distribution parameter can be specified separately. 
This enables the model to deal with various features that can make the task of 
fitting a model very challenging such as censoring, truncation or heavy tails.

@@

A common way of including these covariates is to consider them as regressors 
in a distributional model such that each parameter $\theta$ depends on covariates 
$\bold{Z} = (Z_1,\ldots,Z_m)$. This could for example be done by a linear function 
$g$. In that way the fitted model would be specified as 
$\mathcal{D}(Y, \theta = g(\bold{Z}))$. In order to avoid the restriction to linear 
effects one might also use a set of smooth functions $\{f_i\}_{i = 1,\ldots,p}$ 
to specify a more flexible model 
$\mathcal{D}(Y, \theta = f_1(\bold{Z}) + \ldots + f_p(\bold{Z}))$ applying a GAM structure. 
Another approach to distributional regression is proposed by 
\cite{Klein+Kneib+Lang:2015} where each parameter of a possibly complex 
distribution can be modeled by an additive term consisting of a variety of 
different functional effect types. These include non-linear effects, spatial 
effects, random coefficients and interaction surfaces which extends the 
flexibility compared to parametric models such as GAMLSS modeling only 
smooth effects.

@@

\subsection{Distributional tree}
\label{disttree}
Given the fit of a global distributional model $\mathcal{D}(Y, \bold{\hat{\theta}})$, 
the idea is to use the information provided by this model together with a 
set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ 
to decide whether the data set should be split into subgroups or not. 
In particular, the information provided by the fitted model are the scores 
$$
s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
$$ 
These values express how well the model fits the data regarding each 
parameter. Ideally they should fluctuate randomly around zero, 
similar to residuals in ordinary least squares estimation, but in 
this case there is a score value for each pair of estimated parameter 
$\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ 
and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$. This enables the 
model to be sensitive to changes in each of the parameters which is one 
of the main advantages of distributional trees over other tree models 
considering only the mean of the distribution.

@@

\begin{itemize}
\item A distributional model $\mathcal{D}(Y, \theta)$ is fit to the set of 
observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
parameter $\hat{\theta}$. This parameter can be a multidimensional 
vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
depending on the type of distribution. The estimation is done by 
maximizing the log-likelihood function $\ell(\theta; Y)$, i.e.
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
which is equal to solving
$$
\sum_{i=1}^n s(\theta, y_i) = \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\theta; y_i) = 0
$$
for $\theta$.
